Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-25 manuscript/manuscript.tex
section 3

\subsubsection{Posterior Inference} 
The above prior specification induces closed-form full conditionals that can be efficiently updated as part of a Gibbs sampler outlined below. Additional details, including derivations can be found in the Web Appendix. \textbf{[Think about the best way to organize this section. Maybe see my Bayesian Analysis paper for guidance? We can discuss next week.]}

\paragraph{P\'olya--Gamma Data Augmentation for $z_i$} The sampler begins by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. A random variable $w$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in {\rm I\!R}$ if
$$w \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)},$$
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$  
$$p(\boldsymbol\delta_k|\mathbf{z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ik}^{U_{ik}}(1-\pi_{ik})^{1-U_{ik}}$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$. We note that the sum $\sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $e^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )$$

We can use the quantities to re-express the full conditionals for $\boldsymbol\delta_k$ as
$$p(\boldsymbol\delta_k|\mathbf{Z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}} \right )^{U_{ik}} \left (\frac{1}{1 + e^{\eta_{ik}}} \right )^{1-U_{ik}} = p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(e^{\eta_{ik}})^{U_{ik}}}{1 + e^{\eta_{ik}}}$$

which we note is essentially a logistic regression likelihood. We thus apply this P\'olya--Gamma data augmentation scheme to update each $\boldsymbol\delta_k$ ($k = 1,...,K-1$) one at a time based on the binary indicators $U_{ik}$.

\item Emphasize that PG data augmentation for the multinomial model results in a PG mixture of experts model, which is a computationally efficient way to model edge weights.

@@

\begin{algorithm}
\caption{Gibbs Sampler}
\begin{algorithmic}
\small
\linespread{0.5}
    \State $\mathbf{Define} \ n_{iter};\  n_{burn}; \ K; \ \boldsymbol\theta_{init}; \ \boldsymbol\theta_0$
    \State $n_{sim} := n_{iter} - n_{burn}$
    \State $\boldsymbol\theta := \boldsymbol\theta_{init}$
    \For {$\iota = 1,...,n_{sim}$}
        \State \textsc{I. Conditional Imputation}
          \For {$ i = 1,...,n$}
            \State $\mathbf{Draw} \ \mathbf{y}_i^{miss} \ \text{from} \ N_q(\boldsymbol\mu_i^{miss}, \boldsymbol\Sigma_i^{miss})$
            \State $\mathbf{y}_i := \mathbf{y}_i^{miss} \cup \mathbf{y}_i^{obs}$
         \EndFor
        \State \textsc{II. MSN Regression}
          \For {$ k = 1,...,K$}
            \State $n_k := \sum_{i = 1}^n {1}_{z_i = k}$
            \For {$i_k = 1,...,n_k$}
              \State $\mathbf{Draw}\  t_i \ \text{from} \ N_{[0,\infty)}(a_i,A)$
            \EndFor
            \State $\mathbf{X^*}_k := $ \texttt{cbind}$(\mathbf{X}_k,\mathbf{t}_k)$
            \State $\mathbf{Draw} \ \mathbf{B^*}_k \ \text{from} \ \text{MatNorm}(\mathbf{B}_k,\mathbf{L}_k^{-1},\boldsymbol\Sigma_k)$
            \State $\mathbf{Draw} \ \boldsymbol\Sigma_k \ \text{from} \ \text{InvWish}(\nu_k, \mathbf{V}_k)$
          \EndFor
        \State \textsc{III. Multinomial Logit}
          \For {$i = 1,...,n$}
            \For {$k = 1,...,K$}
              \State $\pi_{ik} := P(z_i = k|\mathbf{w}_i,\boldsymbol\delta_k)$
              \State $p_{ik} := P(\mathbf{y}_i|\boldsymbol\beta_k^{*T} \mathbf{x}^*_i,\boldsymbol\Sigma_k)$
            \EndFor
            \State $\mathbf{p}_{z_i} := \frac{\mathbf{p}_i \circ \boldsymbol\pi_i}{\mathbf{p}_i \cdot \boldsymbol\pi_i}$
            \State $\mathbf{Draw} \ z_i \ \text{from} \ \text{Categorical}(\mathbf{p}_{z_i})$
            \For {$k = 1,...,K-1$}
              \State $\mathbf{Draw} \ \boldsymbol\delta_k \ \text{from} \ N(\mathbf{M},\mathbf{S})$
            \EndFor
          \EndFor
        \State $\boldsymbol\theta := \{\mathbf{B^*}, \boldsymbol\Sigma, \mathbf{Z}, \boldsymbol\delta \}$
        \State $\mathbf{Store} \ \boldsymbol\theta$
	  \EndFor
\end{algorithmic}
\end{algorithm}

@@

@@

@@

\textit{\textbf{Discuss non-ignorable missingness here}}

@@

@@

@@

@@

>>>>

\subsubsection{Posterior Inference} 
The above prior specification induces closed-form full conditionals that can be efficiently updated as part of a Gibbs sampler outlined below. A programatic sketch of our MCMC algorithm is given in table (\ref{alg:MCMC}). Additional details, including derivations can be found in the Web Appendix.

\paragraph{P\'olya--Gamma Data Augmentation for $z_i$} The sampler begins by updating the latent cluster indicators $z_i$ $(i=1,\ldots,n)$ from its multinomial logit full conditional. To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson \textit{et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a P\'olya--Gamma mixture of experts model, a computationally efficient way to model edge weights (\textbf{cite}). A random variable $w$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in {\rm I\!R}$ if
\begin{equation}
w \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson \textit{et al.} (2013) establish that for the $\text{PG}(b,c)$ density, and for $a,\eta \in {\rm I\!R}$,

\begin{equation}
\frac{(e^{\eta})^a}{(1 + e^{\eta})^b} = 2^{-b}e^{\kappa \eta} \int_{0}^{\infty} e^{-w\eta^2/2}p(w|b,c = 0)dw. \label{eq:pg2}
\end{equation}
where $\kappa = a - b/2$. Polson \textit{et al.} also show that the conditional distribution $p(w|b,c)$ results from an ``exponential tilting'' of the $\text{PG}(b,0)$ density, thus

\begin{equation}
	p(w|b,c) = \frac{e^{-c^2w/2}p(w|b,0)}{E_w[e^{-c^2w/2}]} = \frac{e^{-c^2w/2}p(w|b,0)}{\int_0^\infty e^{-c^2w/2}p(w|b,0)dw}. \label{eq:pg3}
\end{equation}

To make use of P\'olya--Gamma data augmentation, we first write the full conditional distribution of $\boldsymbol\delta_k$ as the prior for $\boldsymbol\delta_k$ times the multinomial likelihood for all $z_i$. $$p(\boldsymbol\delta_k|\mathbf{z},\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ik}^{U_{ik}}(1-\pi_{ik})^{1-U_{ik}},$$
where $p(\boldsymbol\delta_k)$ denotes the prior distribution of $\boldsymbol\delta_k$, $U_{ik} = {1}_{z_i = k}$ is an indicator that subject $i$ belongs to cluster $k$, and $\pi_{ik}$ is defined as in Section 3.4. We can rewrite $\pi_{ik}$ as follows
$$\pi_{ik} = P(U_{ik} = 1) = \frac{e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}}{1 + e^{\mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}}} = \frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}}$$
where ${c}_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ and $\eta_{ik} = \mathbf{w}_i^T \boldsymbol\delta_k - {c}_{ik}$. We note that the sum $\sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}}$ includes the reference category, but since we fix $\boldsymbol\delta_K = \mathbf{0}$, we have $e^{\mathbf{w}_i^T \boldsymbol\delta_K} = 1$, and hence
$$c_{ik} = \log \sum_{h \ne k} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} = \log \left ( 1 + \sum_{h \notin \{k,K \}} e^{\mathbf{w}_i^T \boldsymbol\delta_{h}} \right )$$

We can use the quantities to re-express the full conditionals for $\boldsymbol\delta_k$ as
\begin{eqnarray}
	p(\boldsymbol\delta_k|\mathbf{Z},\boldsymbol\delta_{h \ne k}) &\propto& p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \left (\frac{e^{\eta_{ik}}}{1 + e^{\eta_{ik}}} \right )^{U_{ik}} \left (\frac{1}{1 + e^{\eta_{ik}}} \right )^{1-U_{ik}} \nonumber\\ &=& p(\boldsymbol\delta_k) \prod_{i = 1}^n \frac{(e^{\eta_{ik}})^{U_{ik}}}{1 + e^{\eta_{ik}}} \label{eq:pglogit}
\end{eqnarray}
which we note is essentially a logistic regression likelihood. We thus apply this P\'olya--Gamma data augmentation scheme to update each $\boldsymbol\delta_k$ ($k = 1,...,K-1$) one at a time based on the binary indicators $U_{ik}$.

@@

See Table \ref{alg:MCMC}.

\begin{algorithm}
\caption{Gibbs Sampler}
\label{alg:MCMC}
\begin{algorithmic}
\small
\linespread{0.5}
    \State $\mathbf{Define} \ n_{iter};\  n_{burn}; \ K; \ \boldsymbol\theta_{init}; \ \boldsymbol\theta_0$
    \State $n_{sim} := n_{iter} - n_{burn}$
    \State $\boldsymbol\theta := \boldsymbol\theta_{init}$
    \For {$\iota = 1,...,n_{sim}$}
        \State \textsc{I. Conditional Imputation}
          \For {$ i = 1,...,n$}
            \State $\mathbf{Draw} \ \mathbf{y}_i^{miss} \ \text{from} \ N_q(\boldsymbol\mu_i^{miss}, \boldsymbol\Sigma_i^{miss})$
            \State $\mathbf{y}_i := \mathbf{y}_i^{miss} \cup \mathbf{y}_i^{obs}$
         \EndFor
        \State \textsc{II. MSN Regression}
          \For {$ k = 1,...,K$}
            \State $n_k := \sum_{i = 1}^n {1}_{z_i = k}$
            \For {$i_k = 1,...,n_k$}
              \State $\mathbf{Draw}\  t_i \ \text{from} \ N_{[0,\infty)}(a_i,A)$
            \EndFor
            \State $\mathbf{X^*}_k := $ \texttt{cbind}$(\mathbf{X}_k,\mathbf{t}_k)$
            \State $\mathbf{Draw} \ \mathbf{B^*}_k \ \text{from} \ \text{MatNorm}(\mathbf{B}_k,\mathbf{L}_k^{-1},\boldsymbol\Sigma_k)$
            \State $\mathbf{Draw} \ \boldsymbol\Sigma_k \ \text{from} \ \text{InvWish}(\nu_k, \mathbf{V}_k)$
          \EndFor
        \State \textsc{III. Multinomial Logit}
          \For {$i = 1,...,n$}
            \For {$k = 1,...,K$}
              \State $\pi_{ik} := P(z_i = k|\mathbf{w}_i,\boldsymbol\delta_k)$
              \State $p_{ik} := P(\mathbf{y}_i|\boldsymbol\beta_k^{*T} \mathbf{x}^*_i,\boldsymbol\Sigma_k)$
            \EndFor
            \State $\mathbf{p}_{z_i} := \frac{\mathbf{p}_i \circ \boldsymbol\pi_i}{\mathbf{p}_i \cdot \boldsymbol\pi_i}$
            \State $\mathbf{Draw} \ z_i \ \text{from} \ \text{Categorical}(\mathbf{p}_{z_i})$
            \For {$k = 1,...,K-1$}
              \State $\mathbf{Draw} \ \boldsymbol\delta_k \ \text{from} \ N(\mathbf{M},\mathbf{S})$
            \EndFor
          \EndFor
        \State $\boldsymbol\theta := \{\mathbf{B^*}, \boldsymbol\Sigma, \mathbf{Z}, \boldsymbol\delta \}$
        \State $\mathbf{Store} \ \boldsymbol\theta$
	  \EndFor
\end{algorithmic}
\end{algorithm}

@@

We monitor conver convergence of our MCMC algorithm through the use of standard approaches such Geweke's (1992) Z-diagnostic. In simulation studies under realistic parameter settings, we observed relatively fast convergence of all MCMC chains (i.e. within 1,000 iterations).

@@

A problematic feature of mixture models such as the one proposed here is that the full model likelihood is invariant to permutations of $\mathbf{z} = \{ z_1, z_2,..., z_n \}$, the cluster labeling of each observation, the result of which is label switching across MCMC iterations - a process in which draws of cluster-specific parameters may be assigned to the wrong cluster at some point during the MCMC simulation. After conducting simulation studies under a wide variety of realistic parameter settings, as detailed in Section 4, we found little evidence of label switching in posterior draws of $\mathbf{z}$. When label switching was observed, we implemented \textit{post hoc} relabling algorithms as described in Panagiotis 2016 and implemented in the \texttt{label.switching} package in \texttt{R}. We discuss other possible label switching remedies in Section 6.

@@

@@

\item Discuss other label switching approaches

\item Discuss skew-t?

@@

\bibitem{ } Geweke, J. Evaluating the accuracy of sampling-based ap- proaches to calculating posterior moments. \textit{Bayesian Statistics} Vol. 4, J. M. Bernardo, J. Berger, A. P. Dawid, and A.F.M. Smith (eds), 169–193. 1992. Cambridge, U.K.: Oxford University Press.

@@

\bibitem{ } Panagiotis Papastamoulis (2016). label.switching: An R Package for Dealing with the Label Switching Problem in MCMC Outputs. \textit{Journal of Statistical Software}, 69(1), 1-24. doi:10.18637/jss.v069.c01

@@

\bibitem{ } Watanabe S. Asymptotic equivalence of Bayes cross validation and widely applicable information criterion in singular learning theory. \textit{Journal of Machine Learning Research}. 2010;11(Dec):3571-94.
